---
title: "05_Reproducible_Research_Course_Project_2"
author: "Alexander Cormack"
date: "2023-05-20"
output: html_document
---

# Project description

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

* [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb]

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC Storm Events-FAQ Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

## Review criteria

1.	Has either a (1) valid RPubs URL pointing to a data analysis document for this assignment been submitted; or (2) a complete PDF file presenting the data analysis been uploaded?
2.	Is the document written in English?
3.	Does the analysis include description and justification for any data transformations?
4.	Does the document have a title that briefly summarizes the data analysis?
5.	Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?
6.	Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?
7.	Is there a section titled "Results" where the main results are presented?
8.	Is there at least one figure in the document that contains a plot?
9.	Are there at most 3 figures in this document?
10.	Does the analysis start from the raw data file (i.e. the original .csv.bz2 file)?
11.	Does the analysis address the question of which types of events are most harmful to population health?
12.	Does the analysis address the question of which types of events have the greatest economic consequences?
13.	Do all the results of the analysis (i.e. figures, tables, numerical summaries) appear to be reproducible?
14.	Do the figure(s) have descriptive captions (i.e. there is a description near the figure of what is happening in the figure)?
15.	As far as you can determine, does it appear that the work submitted for this project is the work of the student who submitted it?


## Assignment

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

### Questions

Your data analysis must address the following questions:

1.	Across the United States, which types of events (as indicated in the <span style="color: red;">EVTYPE</span> variable) are most harmful with respect to population health?
2.	Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

### Requirements

For this assignment you will need some specific tools

* RStudio: You will need RStudio to publish your completed analysis document to RPubs. You can also use RStudio to edit/write your analysis.
* knitr: You will need the knitr package in order to compile your R Markdown document and convert it to HTML

## Document Layout

* Language: Your document should be written in English.
* Title: Your document should have a title that briefly summarizes your data analysis
* Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.
* There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the <span style="color: red;">cache = TRUE</span> option for certain code chunks.
* There should be a section titled Results in which your results are presented.
* You may have other sections in your analysis, but Data Processing and Results are required.
* The analysis document must have at least one figure containing a plot.
* Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.
* You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that <span style="color: red;">echo = TRUE</span> for every code chunk (this is the default setting in knitr).

## Publishing Your Analysis

For this assignment you will need to publish your analysis on RPubs.com. If you do not already have an account, then you will have to create a new account. After you have completed writing your analysis in RStudio, you can publish it to RPubs by doing the following:

1.	In RStudio, make sure your R Markdown document (<span style="color: red;">.Rmd</span>) document is loaded in the editor
2.	Click the <span style="color: red;">Knit HTML</span> button in the doc toolbar to preview your document.
3.	In the preview window, click the <span style="color: red;">Publish</span> button.

Once your document is published to RPubs, you should get a unique URL to that document. Make a note of this URL as you will need it to submit your assignment.

NOTE: If you are having trouble connecting with RPubs due to proxy-related or other issues, you can upload your final analysis document file as a PDF to Coursera instead.

## Submitting Your Assignment

In order to submit this assignment, you must copy the RPubs URL for your completed data analysis document in to the peer assessment question.
If you choose to submit as a PDF, please insert an obvious placeholder URL (e.g. https://google.com) in order to allow submission.


# Title




# Synopsis






# Data Processing

As a first step the code chunks options in this R markdowwn document are set to echo = TRUE.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

The libraries that we will be used for this report are also loaded.

```{r load libraries, message = FALSE}
library(dplyr)
# library(ggplot2)
# library(lubridate)
```



Here  the contents of the zip file are downloaded and read into a data frame. N.B. the 'read.csv' command can handle compressed files automatically.

```{r download and read file, cache = TRUE, message = FALSE}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
destfile <- "./data/05_assignment2_data.csv.bz2"
download.file(url, destfile)
storm_data <- read.csv("./data/05_assignment2_data.csv.bz2")
```

The data processing begins with an analysis of the structure of the data.

```{r data structure}
str(storm_data)
```

There are 902297 rows and 37 columns. The task at hand is the following:

1.	Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2.	Across the United States, which types of events have the greatest economic consequences?

So the columns of interest are EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG and CROPDMGEXP.

First up is to check for anomalies (e.g. duplicate values) in the EVTYPE column.

```{r check EVTYPE column, results = "hide"}
sort(unique(storm_data$EVTYPE))
```

The National Weather Service Storm Data Documentation lists 48 event types, whereas the data set has 985 unique values. For the sake of space the results of the above code are not displayed. However, they shohw that some of the causes include extra white space, lowercase/uppercase duplicates, sub-classifications (e.g. WET MICROBURST should be classified as THUNDERSTORM WIND), spelling errors or outright anomalies (see for example the many instances of "Summary").

The code below deals with the duplicates.

```{r remove duplicates, cache = TRUE, message = FALSE}
storm_data$EVTYPE <- trimws(storm_data$EVTYPE) ## trim leading and trailing white space
storm_data$EVTYPE <- gsub("\\s+"," ", storm_data$EVTYPE)  ## remove extra whites pace between words
storm_data$EVTYPE <- toupper(storm_data$EVTYPE)  ## convert all strings to uppercase


## The following code reclassifies many of the event types into one of the 48 categories. Misspellings are also taken into account

avalanche <- c("AVALANCH")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% avalanche, "AVALANCHE"))

blizzard <- c("BLIZZARD AND EXTREME WIND CHIL", "BLIZZARD AND HEAVY SNOW", "BLIZZARD SUMMARY", 
                     "BLIZZARD WEATHER","BLIZZARD/FREEZING RAIN", "BLIZZARD/HEAVY SNOW", "BLIZZARD/HIGH WIND", 
                     "BLIZZARD/WINTER STORM", "BLOWING SNOW", "BLOWING SNOW- EXTREME WIND CHI", 
                     "BLOWING SNOW & EXTREME WIND CH", "BLOWING SNOW/EXTREME WIND CHIL", "GROUND BLIZZARD")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% blizzard, "BLIZZARD"))


coastal_flood <- c("ASTRONOMICAL HIGH TIDE", "BEACH EROSIN", "BEACH EROSION", "BEACH EROSION/COASTAL FLOOD", 
                      "BEACH FLOOD", "BLOW-OUT TIDE", "BLOW-OUT TIDES", "BREAKUP FLOODING", "COASTAL EROSION", 
                      "COASTAL FLOODING", "COASTAL FLOODING/EROSION", "COASTAL/TIDAL FLOOD", "COASTALFLOOD", 
                      "CSTL FLOODING/EROSION", "EROSION/CSTL FLOOD", "HIGH TIDES", "HIGH WATER")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% coastal_flood, "COASTAL FLOOD"))

cold_wind_chill <- c("COLD", "COLD AND FROST", "COLD AND SNOW", "COLD AND WET CONDITIONS", "COLD TEMPERATURE", 
                        "COLD TEMPERATURES", "COLD WAVE", "COLD WEATHER", "COLD WIND CHILL TEMPERATURES", 
                        "COLD/WINDS", "COOL AND WET", "COOL SPELL", "EXTENDED COLD")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% cold_wind_chill, "COLD/WIND CHILL"))


dense_fog <- c("FOG", "FOG AND COLD TEMPERATURES")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% dense_fog, "FOG"))

drought <- c("ABNORMALLY DRY", "BELOW NORMAL PRECIPITATION", "DRIEST MONTH", "DROUGHT/EXCESSIVE HEAT", "DRY", 
                "DRY CONDITIONS", "DRY HOT WEATHER", "DRY PATTERN", "DRY SPELL", "DRY WEATHER", "DRYNESS", 
                "EXCESSIVELY DRY")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% drought, "DROUGHT"))

dust_devil <- c("DUST DEVEL", "DUST DEVIL WATERSPOUT")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% dust_devil, "DUST DEVIL"))

dust_storm <- c("BLOWING DUST", "DUST STORM/HIGH WINDS", "DUSTSTORM")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% drought, "DUST STORM"))

excessive_heat <- c("EXCESSIVE", "EXCESSIVE HEAT/DROUGHT", "EXTREME HEAT")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% excessive_heat, "EXCESSIVE HEAT"))

extreme_cold <- c("BITTER WIND CHILL", "BITTER WIND CHILL TEMPERATURES", "EXCESSIVE COLD", "EXTREME COLD", 
                  "EXTREME WIND CHILL", "EXTREME WIND CHILL/BLOWING SNO", "EXTREME WIND CHILLS", "EXTREME WINDCHILL", 
                  "EXTREME WINDCHILL TEMPERATURES", "EXTREME/RECORD COLD", "HYPERTHERMIA/EXPOSURE", "HYPOTHERMIA", 
                  "HYPOTHERMIA/EXPOSURE")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% extreme_cold, "EXTREME COLD/WIND CHILL"))

flash_flood <- c("DAM BREAK", "DAM FAILURE", "FLASH FLOOD - HEAVY RAIN", "FLASH FLOOD FROM ICE JAMS", 
                    "FLASH FLOOD LANDSLIDES", "FLASH FLOOD WINDS", "FLASH FLOOD/", "FLASH FLOOD/ FLOOD", 
                    "FLASH FLOOD/ STREET", "FLASH FLOOD/FLOOD", "FLASH FLOOD/HEAVY RAIN", "FLASH FLOOD/LANDSLIDE", 
                    "FLASH FLOODING", "FLASH FLOODING/FLOOD", "FLASH FLOODING/THUNDERSTORM WI", "FLASH FLOODS", 
                    "FLASH FLOOODING", "FLOOD FLASH", "FLOOD FLOOD/FLASH", "FLOOD/FLASH", "FLOOD/FLASH FLOOD", 
                    "FLOOD/FLASH FLOODING", "FLOOD/FLASH/FLOOD", "FLOOD/FLASHFLOOD", "HIGHWAY FLOODING")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% flash_flood, "FLASH FLOOD"))

flood <- c("FLOOD & HEAVY RAIN", "FLOOD WATCH/", "FLOOD/RAIN/WIND", "FLOOD/RAIN/WINDS", "FLOOD/RIVER FLOOD", 
              "FLOOD/STRONG WIND", "FLOODING", "FLOODING/HEAVY RAIN", "FLOODS")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% flood, "FLOOD"))

freezing_fog <- c("BLACK ICE", "FREEZE", "GLAZE", "GLAZE ICE", "GLAZE/ICE STORM")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% freezing_fog, "FREEZING FOG"))

frost_freeze <- c("AGRICULTURAL FREEZE", "DAMAGING FREEZE", "EARLY FREEZE", "EARLY FROST", "FIRST FROST", "FROST", 
                     "FROST\\FREEZE", "HARD FREEZE")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% frost_freeze, "FROST/FREEZE"))

funnel_cloud <- c("COLD AIR FUNNEL", "COLD AIR FUNNELS", "FUNNEL", "FUNNEL CLOUD.", "FUNNEL CLOUD/HAIL", "FUNNEL CLOUDS", 
                     "FUNNELS")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% funnel_cloud, "FUNNEL CLOUD"))

hail <- c("DEEP HAIL", "HAIL 0.75", "HAIL 0.88", "HAIL 075", "HAIL 088", "HAIL 1.00", "HAIL 1.75", "HAIL 1.75)", 
             "HAIL 100", "HAIL 125", "HAIL 150", "HAIL 175", "HAIL 200", "HAIL 225", "HAIL 275", "HAIL 450", "HAIL 75", 
             "HAIL 80", "HAIL 88", "HAIL ALOFT", "HAIL DAMAGE", "HAIL FLOODING", "HAIL STORM", "HAILSTORMS")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% hail, "HAIL"))

heat <- c("ABNORMAL WARMTH", "HEAT DROUGHT", "HEAT WAVE", "HEAT WAVE DROUGHT", "HEAT WAVES", "HEAT/DROUGHT", "HEATBURST", 
             "HIGH TEMPERATURE RECORD", "HOT AND DRY", "HOT PATTERN", "HOT SPELL", "HOT WEATHER", "HOT/DRY PATTERN")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% heat, "HEAT"))

heavy_rain <- c("ABNORMALLY WET", "EARLY RAIN", "EXCESSIVE PRECIPITATION", "EXCESSIVE RAIN", "EXCESSIVE RAINFALL", 
                   "EXCESSIVE WETNESS", "EXTREMELY WET", "HEAVY MIX", "HEAVY PRECIPATATION", "HEAVY PRECIPITATION", 
                   "HEAVY RAIN", "HEAVY RAIN AND FLOOD", "HEAVY RAIN AND WIND", "HEAVY RAIN EFFECTS", "HEAVY RAIN/FLOODING", 
                   "HEAVY RAIN/HIGH SURF", "HEAVY RAIN/LIGHTNING", "HEAVY RAIN/MUDSLIDES/FLOOD", "HEAVY RAIN/SEVERE WEATHER", 
                   "HEAVY RAIN/SMALL STREAM URBAN", "HEAVY RAIN/SNOW", "HEAVY RAIN/URBAN FLOOD", "HEAVY RAIN/WIND", 
                   "HEAVY RAIN; URBAN FLOOD WINDS;", "HEAVY RAINFALL", "HEAVY RAINS", "HEAVY RAINS/FLOODING", 
                   "HEAVY SHOWER", "HEAVY SHOWERS", "HVY RAIN")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% heavy_rain, "HEAVY RAIN"))

heavy_snow <- c("ACCUMULTATED SNOWFALL", "EARLY SNOW", "EARLY SNOWFALL", "EXCESSIVE SNOW", "FALLING SNOW/ICE", 
                   "FIRST SNOW", "HEAVY LAKE SNOW", "HEAVY SNOW-SQUALLS", "HEAVY SNOW & ICE", "HEAVY SNOW AND", 
                   "HEAVY SNOW AND HIGH WINDS", "HEAVY SNOW AND ICE", "HEAVY SNOW AND ICE STORM", "HEAVY SNOW AND STRONG WINDS", 
                   "HEAVY SNOW ANDBLOWING SNOW", "HEAVY SNOW FREEZING RAIN", "HEAVY SNOW SHOWER", "HEAVY SNOW SQUALLS", 
                   "HEAVY SNOW/BLIZZARD", "HEAVY SNOW/BLIZZARD/AVALANCHE", "HEAVY SNOW/BLOWING SNOW", "HEAVY SNOW/FREEZING RAIN", 
                   "HEAVY SNOW/HIGH", "HEAVY SNOW/HIGH WIND", "HEAVY SNOW/HIGH WINDS", "HEAVY SNOW/HIGH WINDS & FLOOD", 
                   "HEAVY SNOW/HIGH WINDS/FREEZING", "HEAVY SNOW/ICE", "HEAVY SNOW/ICE STORM", "HEAVY SNOW/SLEET", 
                   "HEAVY SNOW/SQUALLS", "HEAVY SNOW/WIND", "HEAVY SNOW/WINTER STORM", "HEAVY SNOWPACK")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% heavy_snow, "HEAVY SNOW"))

high_surf <- c("HAZARDOUS SURF", "HEAVY SEAS", "HEAVY SURF", "HEAVY SURF AND WIND", "HEAVY SURF COASTAL FLOODING", 
                  "HEAVY SURF/HIGH SURF", "HEAVY SWELLS", "HEAVY WET SNOW", "HIGH", "HIGH SEAS", "HIGH SURF", 
                  "HIGH SURF ADVISORIES", "HIGH SURF ADVISORY", "HIGH SWELLS", "HIGH WAVES")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% high_surf, "HIGH SURF"))

high_wind <- c("HIGH WIND (G40)", "HIGH WIND 48", "HIGH WIND 63", "HIGH WIND 70", "HIGH WIND AND HEAVY SNOW", 
                  "HIGH WIND AND HIGH TIDES", "HIGH WIND AND SEAS", "HIGH WIND DAMAGE", "HIGH WIND/ BLIZZARD", 
                  "HIGH WIND/BLIZZARD", "HIGH WIND/BLIZZARD/FREEZING RA", "HIGH WIND/HEAVY SNOW", "HIGH WIND/LOW WIND CHILL", 
                  "HIGH WIND/SEAS", "HIGH WIND/WIND CHILL", "HIGH WIND/WIND CHILL/BLIZZARD", "HIGH WINDS", 
                  "HIGH WINDS 55", "HIGH WINDS 57", "HIGH WINDS 58", "HIGH WINDS 63", "HIGH WINDS 66", "HIGH WINDS 67", 
                  "HIGH WINDS 73", "HIGH WINDS 76", "HIGH WINDS 80", "HIGH WINDS 82", "HIGH WINDS AND WIND CHILL", 
                  "HIGH WINDS DUST STORM", "HIGH WINDS HEAVY RAINS", "HIGH WINDS/", "HIGH WINDS/COASTAL FLOOD", 
                  "HIGH WINDS/COLD", "HIGH WINDS/FLOODING", "HIGH WINDS/HEAVY RAIN", "HIGH WINDS/SNOW")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% high_wind, "HIGH WIND"))

hurricane <- c("HURRICANE-GENERATED SWELLS", "HURRICANE EDOUARD", "HURRICANE EMILY", "HURRICANE ERIN", 
                  "HURRICANE FELIX", "HURRICANE GORDON", "HURRICANE OPAL", "HURRICANE OPAL/HIGH WINDS", "HURRICANE/TYPHOON")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% hurricane, "HURRICANE"))

sleet <- c("FREEZING DRIZZLE", "FREEZING DRIZZLE AND FREEZING", "FREEZING RAIN", "FREEZING RAIN AND SLEET", 
              "FREEZING RAIN AND SNOW", "FREEZING RAIN SLEET AND", "FREEZING RAIN SLEET AND LIGHT", "FREEZING RAIN/SLEET", 
              "FREEZING RAIN/SNOW", "FREEZING SPRAY")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% sleet, "SLEET"))

strong_wind <- c("GUSTY LAKE WIND", "GUSTY WIND", "GUSTY WIND/HAIL", "GUSTY WIND/HVY RAIN", "GUSTY WIND/RAIN", 
                    "GUSTY WINDS")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% strong_wind, "STRONG WIND"))

thunderstorm_wind <- c("DOWNBURST", "DOWNBURST WINDS", "DRY MICROBURST", "DRY MICROBURST 50", "DRY MICROBURST 53", 
                          "DRY MICROBURST 58", "DRY MICROBURST 61", "DRY MICROBURST 84", "DRY MICROBURST WINDS", 
                          "DRY MIRCOBURST WINDS",  "GUSTNADO", "GUSTNADO AND", "GUSTY THUNDERSTORM WIND", 
                          "GUSTY THUNDERSTORM WINDS", "THUDERSTORM WINDS", "THUNDEERSTORM WINDS", "THUNDERESTORM WINDS", 
                          "THUNDERSNOW", "THUNDERSNOW SHOWER", "THUNDERSTORM", "THUNDERSTORM DAMAGE", "THUNDERSTORM DAMAGE TO", 
                          "THUNDERSTORM HAIL", "THUNDERSTORM W INDS", "THUNDERSTORM WIND (G40)", "THUNDERSTORM WIND 50", 
                          "THUNDERSTORM WIND 52", "THUNDERSTORM WIND 56", "THUNDERSTORM WIND 59", "THUNDERSTORM WIND 59 MPH", 
                          "THUNDERSTORM WIND 59 MPH.", "THUNDERSTORM WIND 60 MPH", "THUNDERSTORM WIND 65 MPH", 
                          "THUNDERSTORM WIND 65MPH", "THUNDERSTORM WIND 69", "THUNDERSTORM WIND 98 MPH", "THUNDERSTORM WIND G50", 
                          "THUNDERSTORM WIND G51", "THUNDERSTORM WIND G52", "THUNDERSTORM WIND G55", "THUNDERSTORM WIND G60", 
                          "THUNDERSTORM WIND G61", "THUNDERSTORM WIND TREES", "THUNDERSTORM WIND.", "THUNDERSTORM WIND/ TREE", 
                          "THUNDERSTORM WIND/ TREES", "THUNDERSTORM WIND/AWNING", "THUNDERSTORM WIND/HAIL", 
                          "THUNDERSTORM WIND/LIGHTNING", "THUNDERSTORM WINDS", "THUNDERSTORM WINDS 13", "THUNDERSTORM WINDS 2", 
                          "THUNDERSTORM WINDS 50", "THUNDERSTORM WINDS 52", "THUNDERSTORM WINDS 53", "THUNDERSTORM WINDS 60", 
                          "THUNDERSTORM WINDS 61", "THUNDERSTORM WINDS 62", "THUNDERSTORM WINDS 63 MPH", 
                          "THUNDERSTORM WINDS AND", "THUNDERSTORM WINDS FUNNEL CLOU", "THUNDERSTORM WINDS G", 
                          "THUNDERSTORM WINDS G60", "THUNDERSTORM WINDS HAIL", "THUNDERSTORM WINDS HEAVY RAIN", 
                          "THUNDERSTORM WINDS LE CEN", "THUNDERSTORM WINDS LIGHTNING", "THUNDERSTORM WINDS SMALL STREA", 
                          "THUNDERSTORM WINDS URBAN FLOOD", "THUNDERSTORM WINDS.", "THUNDERSTORM WINDS/ FLOOD", 
                          "THUNDERSTORM WINDS/ HAIL", "THUNDERSTORM WINDS/FLASH FLOOD", "THUNDERSTORM WINDS/FLOODING", 
                          "THUNDERSTORM WINDS/FUNNEL CLOU", "THUNDERSTORM WINDS/HAIL", "THUNDERSTORM WINDS/HEAVY RAIN", 
                          "THUNDERSTORM WINDS53", "THUNDERSTORM WINDSHAIL", "THUNDERSTORM WINDSS", "THUNDERSTORM WINS", 
                          "THUNDERSTORMS", "THUNDERSTORMS WIND", "THUNDERSTORMS WINDS", "THUNDERSTORMW", "THUNDERSTORMW 50", 
                          "THUNDERSTORMW WINDS", "THUNDERSTORMWINDS", "THUNDERSTROM WIND", "THUNDERSTROM WINDS", 
                          "THUNDERTORM WINDS", "THUNDERTSORM WIND", "THUNDESTORM WINDS", "THUNERSTORM WINDS")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% thunderstorm_wind, "THUNDERSTORM WIND"))

tornado <- c("COLD AIR TORNADO")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% tornado, "TORNADO"))

tropical_storm <- c("COASTAL STORM", "COASTAL SURGE", "COASTALSTORM", "GRADIENT WIND", "GRADIENT WINDS")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% tropical_storm, "TROPICAL STORM"))

wildfire <- c("BRUSH FIRE", "BRUSH FIRES", "FOREST FIRES", "GRASS FIRES")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% wildfire, "WILDFIRE"))

winter_weather <- c("DRIFTING SNOW")
storm_data$EVTYPE <- sapply(storm_data$EVTYPE, function(x) replace(x, x %in% winter_weather, "WINTER WEATHER"))


## The following code drops rows from the dataframe where the event type value is truly anomalous

anomalous_values <- c("?", "APACHE COUNTY", "DROWNING", "SUMMARY AUGUST 10", "SUMMARY AUGUST 11", "SUMMARY AUGUST 17", 
                      "SUMMARY AUGUST 2-3", "SUMMARY AUGUST 21", "SUMMARY AUGUST 28", "SUMMARY AUGUST 4", 
                      "SUMMARY AUGUST 7", "SUMMARY AUGUST 9", "SUMMARY JAN 17", "SUMMARY JULY 23-24", "SUMMARY JUNE 18-19", 
                      "SUMMARY JUNE 5-6", "SUMMARY JUNE 6", "SUMMARY OF APRIL 12", "SUMMARY OF APRIL 13", 
                      "SUMMARY OF APRIL 21", "SUMMARY OF APRIL 27","SUMMARY OF APRIL 3RD", "SUMMARY OF AUGUST 1", 
                      "SUMMARY OF JULY 11", "SUMMARY OF JULY 2", "SUMMARY OF JULY 22", "SUMMARY OF JULY 26", 
                      "SUMMARY OF JULY 29", "SUMMARY OF JULY 3", "SUMMARY OF JUNE 10", "SUMMARY OF JUNE 11", 
                      "SUMMARY OF JUNE 12", "SUMMARY OF JUNE 13", "SUMMARY OF JUNE 15", "SUMMARY OF JUNE 16", 
                      "SUMMARY OF JUNE 18", "SUMMARY OF JUNE 23", "SUMMARY OF JUNE 24", "SUMMARY OF JUNE 3", 
                      "SUMMARY OF JUNE 30", "SUMMARY OF JUNE 4", "SUMMARY OF JUNE 6", "SUMMARY OF MARCH 14", 
                      "SUMMARY OF MARCH 23", "SUMMARY OF MARCH 24", "SUMMARY OF MARCH 24-25", "SUMMARY OF MARCH 27", 
                      "SUMMARY OF MARCH 29", "SUMMARY OF MAY 10", "SUMMARY OF MAY 13", "SUMMARY OF MAY 14", 
                      "SUMMARY OF MAY 22", "SUMMARY OF MAY 22 AM", "SUMMARY OF MAY 22 PM", "SUMMARY OF MAY 26 AM", 
                      "SUMMARY OF MAY 26 PM", "SUMMARY OF MAY 31 AM", "SUMMARY OF MAY 31 PM", "SUMMARY OF MAY 9-10", 
                      "SUMMARY SEPT. 25-26", "SUMMARY SEPTEMBER 20", "SUMMARY SEPTEMBER 23", "SUMMARY SEPTEMBER 3", 
                      "SUMMARY SEPTEMBER 4", "SUMMARY: NOV. 16", "SUMMARY: NOV. 6-7", "SUMMARY: OCT. 20-21", 
                      "SUMMARY: OCTOBER 31", "SUMMARY: SEPT. 18")
storm_data <- subset(storm_data, !(EVTYPE %in% anomalous_values))


sort(unique(storm_data$EVTYPE))
```




According to the National Weather Service Storm Data Documentation, property damage estimates *should be rounded to three significant digits, followed by an alphabetical character signifying the magnitude of the number, i.e., 1.55B for $1,550,000,000. Alphabetical characters used to signify magnitude include “K” for thousands, “M” for millions, and “B” for billions.* 

This means the PROPDMG and PROPDMGEXP columns will need to be combined to get numeric values that calculations can be performed on. But first, let's examine the unique values in the PROPDMGEXP column.

```{r check PROPDMGEXO column}
table(storm_data$PROPDMGEXP)
```

As we can see, as well as the magnitude values of "K", "M" and "B" there are 15 other anomalous values. These can be put down to data input errors. Assuming the lower case 'm' should really be an uppercase 'M', it is changed here.

```{r update PROPDMGEXP column}
storm_data$PROPDMGEXP[storm_data$PROPDMGEXP == "m"] <- "M"
```


Since the remaining anomalies only make 0.07% (321 of the 465934) of the records in the column they can be safely ignored.

Here the values in the PROPDMG column are multiplied by the exponent in the PROPDMGEXP column.

```{r multiply PROPDMG by PROPDMGEXP}
storm_data$PROPDMG = with(storm_data, ifelse(PROPDMGEXP == "K", PROPDMG * 1000,
                            ifelse(PROPDMGEXP =="M", PROPDMG * 1000000,
                                   ifelse(PROPDMGEXP == "B", PROPDMG * 1000000000,
                                          PROPDMG * 1))))
```


Turning now to crop damage, here the CROPDMGEXP column is checked for anomalies similar to those in the PROPDMGEXP column.

```{r check CROPDMGEXO column}
table(storm_data$CROPDMGEXP)
```

Again, assuming the lowercase "k" and "m" should be upper case they are changed here.

```{r update CROPDMGEXP column}
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "k"] <- "K"
storm_data$CROPDMGEXP[storm_data$CROPDMGEXP == "m"] <- "M"
```

Here the values in the CROPDMG column are multiplied by the exponent in the CROPDMGEXP column.

```{r multiply CROPDMG by CROPDMGEXP}
storm_data$CROPDMG = with(storm_data, ifelse(CROPDMGEXP == "K", CROPDMG * 1000,
                            ifelse(CROPDMGEXP =="M", CROPDMG * 1000000,
                                   ifelse(CROPDMGEXP == "B", CROPDMG * 1000000000,
                                          CROPDMG * 1))))
```


Here the relevant columns are selected and the data is grouped by event type.

```{r select group and sum}
storm_data %>%
  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, CROPDMG) %>%
  group_by(EVTYPE) %>%
  summarise(
    FATALITIES = sum(FATALITIES),
    INJURIES = sum(INJURIES),
    PROPDMG = sum(PROPDMG),
    CROPDMG = sum(CROPDMG)
  )
```












# Results
















