---
title: "05_Reproducible_Reasearch_Course_Project_1"
author: "Alexander Cormack"
date: "2023-04-15"
output: html_document
keep_md: TRUE
---

# Assignment Instructions
## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

* date: The date on which the measurement was taken in YYYY-MM-DD format

* interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Review criteria

Repo

1.    Valid GitHub URL 

2.    At least one commit beyond the original fork

3.    Valid SHA-1

4.    SHA-1 corresponds to a specific commit

Commit containing full submission

1.    Code for reading in the dataset and/or processing the data

2.    Histogram of the total number of steps taken each day

3.    Mean and median number of steps taken each day

4.    Time series plot of the average number of steps taken

5.    The 5-minute interval that, on average, contains the maximum number of steps

6.    Code to describe and show a strategy for imputing missing data

7.    Histogram of the total number of steps taken each day after missing values are imputed

8.    Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

9.    All of the R code needed to reproduce the results (numbers, plots, etc.) in the report



## Assignment

This assignment will be described in multiple parts. You will need to write a report that answers the questions detailed below. Ultimately, you will need to complete the entire assignment in a **single R markdown** document that can be processed by knitr and be transformed into an HTML file.

Throughout your report make sure you always include the code that you used to generate the output you present. When writing code chunks in the R markdown document, always use echo = TRUE so that someone else will be able to read the code. **This assignment will be evaluated via peer assessment so it is essential that your peer evaluators be able to review the code for your analysis.**

For the plotting aspects of this assignment, feel free to use any plotting system in R (i.e., base, lattice, ggplot2)

Fork/clone the [GitHub repository created for this assignment](http://github.com/rdpeng/RepData_PeerAssessment1). You will submit this assignment by pushing your completed files into your forked repository on GitHub. The assignment submission will consist of the URL to your GitHub repository and the SHA-1 commit ID for your repository state.

NOTE: The GitHub repository also contains the dataset for the assignment so you do not have to download the data separately.

**Loading and preprocessing the data**

Show any code that is needed to

1.    Load the data (i.e. read.csv())

2.    Process/transform the data (if necessary) into a format suitable for your analysis

**What is mean total number of steps taken per day?**

For this part of the assignment, you can ignore the missing values in the dataset.

1.    Calculate the total number of steps taken per day

2.    If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

3.    Calculate and report the mean and median of the total number of steps taken per day

**What is the average daily activity pattern?**

1.    Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2.    Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

**Imputing missing values**

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1.    Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

2.    Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3.    Create a new dataset that is equal to the original dataset but with the missing data filled in.

4.    Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

**Are there differences in activity patterns between weekdays and weekends?**

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1.    Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

2.    Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

## Submitting the assignment

To submit the assignment:

1.    Commit your completed PA1_template.Rmd file to the master branch of your git repository (you should already be on the master branch unless you created new ones)

2.    Commit your PA1_template.md and PA1_template.html files produced by processing your R markdown file with knit2html() function in R (from the knitr package) by running the function from the console.

3.    If your document has figures included (it should) then they should have been placed in the figure/ directory by default (unless you overrided the default). Add and commit the figure/ directory to your git repository so that the figures appear in the markdown file when it displays on github.

4.    Push your master branch to GitHub.

5.    Submit the URL to your GitHub repository for this assignment on the course web site.

In addition to submitting the URL for your GitHub repository, you will need to submit the 40 character SHA-1 hash (as string of numbers from 0-9 and letters from a-f) that identifies the repository commit that contains the version of the files you want to submit. You can do this in GitHub by doing the following

1.    Going to your GitHub repository web page for this assignment

2.    Click on the “?? commits” link where ?? is the number of commits you have in the repository. For example, if you made a total of 10 commits to this repository, the link should say “10 commits”.

3.    You will see a list of commits that you have made to this repository. The most recent commit is at the very top. If this represents the version of the files you want to submit, then just click the “copy to clipboard” button on the right hand side that should appear when you hover over the SHA-1 hash. Paste this SHA-1 hash into the course web site when you submit your assignment. If you don't want to use the most recent commit, then go down and find the commit you want and copy the SHA-1 hash.

A valid submission will look something like (this is just an **example**!)

    1  https://github.com/rdpeng/RepData_PeerAssessment17c376cc54
    2
    3  47f11537f8740af8e07d6facc3d9645




# My Assignment

## Setup

As a first step let's set the code chunks options in this R markdowwn document to echo = TRUE.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's also load all of the libraries that we will be using for this assignment.

```{r load libraries, message = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
```


## Loading and preprocessing the data 

Here we download and list the contents of the zip file.

```{r download and list files}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile <- "./data/05_assignment1_data.zip"
download.file(url, destfile)
unzip("./data/05_assignment1_data.zip", list = TRUE)
```


Next we unzip and read the dataset into an R dataframe

```{r unzip and read, message = FALSE}
activity_df <- read_csv(unzip("./data/05_assignment1_data.zip", "activity.csv", exdir = "./data"))
```

## What is the mean total number of steps taken per day?

Let's have a look at our dataframe.

```{r structure}
str(activity_df)
```
```{r head}
head(activity_df)
```

We see that there are 17,568 rows and three columns: **steps** in numeric format, **date** in date format and **interval** in numeric format. We also appreciate that there are some NA values in the **date** column. 

Now let's calculate the total number of steps taken each day.

The following code groups the data by date and then calculates the total number of steps for each day, producing a dataframe with 61 rows (dates from 1 October 2012 to 30 November 2012) and two columns (dates and steps).

```{r total steps}
steps_day <- activity_df %>% 
        group_by(date) %>% 
        summarise(
        steps = sum(steps)
)
steps_day
```

Now we simply use the summary function to calculate the mean and the median of the total number of steps taken per day.

```{r mean and meadian total steps per day}
summary(steps_day$steps)
```

The mean of 10766 and the median of 10765 show a fairly central distribution of the data.


Let's now see how that looks in the histogram of the total number of steps taken each day.

```{r histogram total steps per day}
ggplot(steps_day, aes(x=steps)) + 
        geom_histogram(bins = 22, color = "black", fill = "steelblue") + 
        ggtitle("Distribution of total steps per day")
```

As we might have expected from the summary of the data, the histogram shows a fairly normal distribution around a total of 10,000 steps per day.



## What is the average daily activity pattern?


The following code groups the data by interval and then calculates the average number of steps for each interval, producing a dataframe with 288 rows (number of 5-minute intervals in each day) and two columns (interval and steps).

Then we find the 5-minute interval with the maximum average number of steps.

```{r average steps per interval}
by_interval <- activity_df %>%
        group_by(interval) %>%
        summarise(
                steps = mean(steps, na.rm = TRUE)
        )
by_interval[which.max(by_interval$steps), ]
```

As we can see the interval with the maximum average number of steps (206 steps) occurred at interval 835, i.e. on average the subject was most active between 8:35 and 8:40 am across the two months in question.


In the time series chart below we can clearly see how the subject was most active on average in the morning  around 8:30 am.

```{r plot average steps per interval}
ggplot(by_interval, aes(x=interval, y=steps)) + 
        geom_line(color = "steelblue") + 
        ggtitle("Average steps per 5-minute interval") + 
        scale_x_continuous(breaks=seq(0,2400,300))
```

## Imputing missing values

First of all, let's calculate the number of missing values in the steps column.

```{r calculate missing values}
sum(is.na(activity_df$steps))
```

Well, that's quite a lot of missing values!

Let's replace them now with the mean value for that 5-minute interval. The code below does just that, first by grouping the data according to interval and then checking for NA values and when found replacing them with the mean number of steps for that interval.

```{r replace missing values}
replaced_missing <- activity_df %>%
        group_by(interval) %>%
        mutate_if(is.numeric,
                 function(x) ifelse(is.na(x),
                                    mean(x, na.rm = TRUE),
                                    x)) 
```

Now let's calculate the mean and median number of steps per day in the new dataframe with the replaced missing values. First we need to ungroup the dataframe (as we had grouped it by interval) and then group it by date. Next we calculate the total number of steps per day and lastly use the summary function to display the mean and median.

```{r mean and median total steps with imputed missing values}
steps_day_no_na <- replaced_missing %>%
        ungroup(interval) %>%
        group_by(date) %>%
        summarise(
        steps = sum(steps)
)
summary(steps_day_no_na$steps)
```

We can immediately see that there is no difference at all to the mean and only a slight difference to the median. We can however observe some difference in the 1st (8841 to 9819) and 3rd (13294 to 12811) quartile figures. So the impact of replacing the missing values with mean values has pushed the data more towards the mean.


Now let's plot the histogram of the total number of steps with the imputed missing values.

```{r histogram total steps with imputed missing values}
ggplot(steps_day_no_na, aes(x=steps)) + 
        geom_histogram(bins = 22, color = "black", fill = "steelblue") + 
        ggtitle("Distribution of total steps per day with imputed missing values")
```

We can see that the data has to a certain extent been 'normalised' with a more even distribution around the centre.


## Are there differences in activity patterns between weekdays and weekends?

First up let's add a column to our dataframe with the imputed missing values to show whether the date is a weekday or a weekend. Using the lubridate function wday() we add a column with the day of the week as a numeric value (1-5 for Monday to Friday, 6-7 for Saturday and Sunday). Then we replace the numeric valued with the string weekday or weekend.

```{r add day of week column}

replaced_missing_wd <- replaced_missing %>%
        mutate(day_of_week = wday(date, week_start = 1)) %>%
        mutate(day_of_week = ifelse(day_of_week < 6, "weekday", "weekend"))
```


Now let's group our data by both interval and the day of the week and calculate the average number of steps per interval.

```{r average steps per interval grouped by day of week}
by_interval_wd <- replaced_missing_wd %>%
        group_by(interval, day_of_week) %>%
        summarise(
                steps = mean(steps, na.rm = TRUE)
        )
```

Lastly let's plot the average steps per 5-minute interval in two panels to compare the weekday activity against the weekend.

```{r plot average steps per interval grouped by day of week}
ggplot(by_interval_wd, aes(x=interval, y=steps, group=day_of_week)) +
        geom_line(color = "steelblue") + 
        ggtitle("Average steps per 5-minute interval") + 
        scale_x_continuous(breaks=seq(0,2400,300)) +
        facet_grid(day_of_week ~ .)
```

From the chart above we can see that the subject is less active early in the morning on the weekends but generally more active throughout the day. We can probably assume that the subject works Monday to Friday in a relatively sedentary job.